---
title: "Covid19"
output:
  pdf_document: default
  html_document: default
date: "2023-02-28"
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```



```{r urls}
url_in<-"https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
filenames<-c(
  "time_series_covid19_confirmed_US.csv",
  "time_series_covid19_confirmed_global.csv",
  "time_series_covid19_deaths_US.csv",
  "time_series_covid19_deaths_global.csv",
  "time_series_covid19_recovered_global.csv"
)
urls<-str_c(url_in,filenames)

```



```{r loads}
us_cases<-read_csv(urls[1])
g_cases<-read_csv(urls[2])
us_death<-read_csv(urls[3])
g_death<-read_csv(urls[4])
g_rec<-read_csv(urls[5])
```

```{r cleanup}
g_cases<-g_cases %>%
  pivot_longer(cols=-c('Province/State','Country/Region', Lat, Long), 
               names_to = "date",values_to = "cases")%>%
  select(-c(Lat,Long))

g_death<-g_death %>%
  pivot_longer(cols=-c('Province/State','Country/Region', Lat, Long), 
               names_to = "date",values_to = "deaths")%>%
  select(-c(Lat,Long))

global <- g_cases %>%
  full_join(g_death) %>%
  rename(Country_Region='Country/Region',
         Province_State='Province/State')%>%
  mutate(date=mdy(date))
global <- global %>%filter(cases>0)

summary(global)


us_cases<-us_cases %>%
  pivot_longer(cols=-(UID:Combined_Key), names_to = "date", values_to = "cases")%>%
  select(Admin2:cases)%>%
  mutate(date=mdy(date))%>%
  select(-c(Lat,Long_))


us_death <-us_death %>%
  pivot_longer(cols=-(UID:Population), names_to = "date", values_to = "deaths")%>%
  select(Admin2:deaths)%>%
  mutate(date=mdy(date))%>%
  select(-c(Lat,Long_))

us <- us_cases %>%
 full_join(us_death)

global<-global %>%
  unite("Combined_Key",
        c(Province_State,Country_Region),
        sep=", ",
        na.rm = TRUE,
        remove=FALSE)
uid_lookup_url<- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid<-read_csv(uid_lookup_url)%>%
 select(-c(Lat,Long_,Combined_Key,code3,iso2,iso3,Admin2))

global <- global %>%
  left_join(uid, by=c("Province_State","Country_Region"))%>%
  select(-c(UID,FIPS)) %>%
  select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)

```



```{r visualize}
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases=sum(cases), deaths= sum(deaths),
    Population = sum(Population))%>%
  mutate(deaths_per_mill=deaths*1000000/Population)%>%
  select(Province_State,Country_Region,date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()

us_totals <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases=sum(cases), deaths= sum(deaths),
    Population = sum(Population))%>%
  mutate(deaths_per_mill=deaths*1000000/Population)%>%
  select(Country_Region,date, cases, deaths, deaths_per_mill, Population)%>%
  ungroup()

us_totals%>%
  ggplot(aes(x=date,y=cases))+
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases"))+
  geom_line(aes(y=deaths,color="deaths"))+
  geom_point(aes(y=deaths,color="deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = "COVID19 in US", y=NULL)
  
us_by_state<-us_by_state%>%
  mutate(new_cases=cases-lag(cases),
         new_deaths=deaths-lag(deaths))
us_totals<-us_totals%>%
  mutate(new_cases=cases-lag(cases),
         new_deaths=deaths-lag(deaths))

us_totals%>%
  ggplot(aes(x=date,y=new_cases))+
  geom_line(aes(color="new_cases"))+
  geom_point(aes(color="new_cases"))+
  geom_line(aes(y=new_deaths,color="new_deaths"))+
  geom_point(aes(y=new_deaths,color="new_deaths"))+
  scale_y_log10()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90))+
  labs(title = "COVID19 in US", y=NULL)

us_by_state<-us_by_state%>%
     mutate(new_cases=cases-lag(cases),
            new_deaths=deaths-lag(deaths))
us_totals<-us_totals%>%
     mutate(new_cases=cases-lag(cases),
            new_deaths=deaths-lag(deaths))

us_totals%>%
     ggplot(aes(x=date,y=new_cases))+
     geom_line(aes(color="new_cases"))+
     geom_point(aes(color="new_cases"))+
     geom_line(aes(y=new_deaths,color="new_deaths"))+
     geom_point(aes(y=new_deaths,color="new_deaths"))+
     scale_y_log10()+
     theme(legend.position = "bottom",
           axis.text.x = element_text(angle=90))+
     labs(title = "COVID19 in US", y=NULL)

us_state_totals<-us_by_state%>%
  group_by(Province_State)%>%
  summarize(deaths=max(deaths),cases=max(cases),
      population=max(Population),
      cases_per_thou=1000*cases/population,
      deaths_per_thou=1000*deaths/population)%>%
  filter(cases>0,population>0)

us_state_totals%>%
  slice_min(deaths_per_thou,n=10)%>%
  select(Province_State,deaths_per_thou,cases_per_thou,everything())
us_state_totals%>%
  slice_max(deaths_per_thou,n=10)%>%
 select(Province_State,deaths_per_thou,cases_per_thou,everything())
```

```{r model}
mod<-lm(deaths_per_thou ~ cases_per_thou,data=us_state_totals)
summary(mod)

us_state_totals %>% slice_max(cases_per_thou)

x_grid<-seq(1,450)
new_df<-tibble(cases_per_thou=x_grid)
us_state_totals_pred<-us_state_totals%>%mutate(pred=predict(mod))
us_state_totals_pred%>% ggplot()+
  geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color="blue")+
  geom_point(aes(x=cases_per_thou, y=pred),color="red")
```

This are my additions:

```{r my_part}
pie(us_state_totals$cases_per_thou, labels = us_state_totals$Province_State, 
    main = "Cases per thousands by state")

pie(us_state_totals$deaths_per_thou, labels = us_state_totals$Province_State, 
    main = "Deaths per thousands by state")


global<-global%>%
     mutate(new_cases=cases-lag(cases),
            new_deaths=deaths-lag(deaths))

global_totals<-global%>%
 group_by(date)%>%
 summarize(deaths=sum(deaths),cases=sum(cases))
```

I know because of the news and because of my understand of how viruses propagate that I was bias towards the belief that small states with many people, such us New York, would be leading the charge in terms of cases. However I very surprise to see that states that have very little population and are relatively big like Wyoming and Alaska have proportionately to their population so many cases. I must also note that after seen the number of cases I am not surprised to see Wyoming also has a high death count because it is my belief and my bias that young people tend to leave that state. This is also probes to me that the measures taken by states like New York were propagation is more likely were quite good.


```{r second_model}
mod2<-lm(deaths ~ date,data=global_totals)
global_pred<-global_totals%>%mutate(pred=predict(mod2))
global_pred%>% ggplot()+
  geom_point(aes(x=date, y=deaths), color="blue")+
  geom_point(aes(x=date, y=pred),color="red")+
  theme(legend.position = "bottom",
           axis.text.x = element_text(angle=90))+
     labs(title = "Cummulative deaths of COVID19 in the world", y=NULL)
```

Regarding the global data I'm just happy that the number of deaths is barely growing, I was biased towards that belief but the graph seems to show undeniably that.